What to simulate? Inferring the right direction for mental rotation

### 2014-04-30 man/cogsci-2014/cogsci-2014.tex
Use posterior odds ratio rather than likelihood ratio

We denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is generated
by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$, where
$\theta$ is a rotation, $\hi$ is the hypothesis that the images depict
the same object, and $\hf$ is the hypothesis that the images depict
mirror-image objects. The posterior probability of each hypothesis
given the observed shapes is then:
$p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta,
h)p(h)p(\theta)\dif\theta$,
where $p(X_b\ \vert\ X_a, \theta, h)$ is the probability that $X_b$
was generated from $X_a$. Because we want to determine which
hypothesis is more likely, the quantity of interest is a ratio
$\ell:=p(\hi\ \vert\ X_a, X_b) / p(\hf\ \vert\ X_a, X_b)$ which
(assuming that all rotations are equally likely) is equivalent to:
\begin{equation}
  \ell = \frac{\left(\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta\right)\cdot{}p_0}{\left(\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta\right)\cdot{}p_1},
  \label{eq:lh-ratio}
\end{equation}
where $p_0=p(h=0)$ and $p_1=p(h=1)$, for brevity. If $\ell > 1$, then
we accept the hypothesis that the images depict the same object
($\hi$); if $\ell < 1$, then we accept the hypothesis that the images
depict flipped objects ($\hf$).

@@

To summarize, we approximate the likelihood term of
Equation~\ref{eq:lh-ratio} using the similarity function defined in
Equation~\ref{eq:similarity}. Because we assume mental rotations are
performed sequentially, this similarity can only be computed for the
actions listed in Equation~\ref{eq:actions}.

@@

In order to approximate Equation~\ref{eq:lh-ratio} using samples of
the similarity function, we must decide \textit{which} places to
sample and \textit{when} stop sampling. The models below differ in how
they make these decisions.

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This gives us the true value of $\theta$, so
Equation~\ref{eq:lh-ratio} becomes a generalized likelihood ratio
test, where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)\cdot{}p_0}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)\cdot{}p_1}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

While the previous few models all focused on \textit{searching} for
the global maximum, we need only \textit{approximate}
Equation~\ref{eq:lh-ratio}. We hypothesize a model based on the idea
of \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu}: instead of
searching for a maximum, we maintain a probability distribution over
our \textit{estimate} of Equation~\ref{eq:lh-ratio}, and then sample
actions which are expected to improve that estimate.  This strategy
has the benefits that it does not make assumptions about the scale of
the similarity function; and, by choosing to sample places which are
informative, this method implicitly minimizes the amount of rotation.

@@

Assuming independence, we can now write
$p(Z_h)\approx\mathcal{N}(Z_h\ \vert\ \mathbb{E}[Z_h],
\mathbb{V}(Z_h))$,
which gives us a distribution over the likelihood ratio in
Equation~\ref{eq:lh-ratio}:
$p(\ell)\approx\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0],
\mathbb{V}(Z_0))\cdot{}p_0/\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1],
\mathbb{V}(Z_1))\cdot{}p_1$.
This distribution cannot easily be calculated, but we are only
interested in whether $Z_0>Z_1$ or $Z_1>Z_0$. So, we use
$Z_D=p_0\cdot{}Z_0-p_1\cdot{}Z_1$ and compute
$p(Z_D)\propto\mathcal{N}(p_0\cdot{}\mathbb{E}[Z_0] -
p_1\cdot{}\mathbb{E}[Z_1], p_0^2\cdot{}\mathbb{V}(Z_0) +
p_1^2\cdot{}\mathbb{V}(Z_1))$.
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$\hi$, and when $p(Z_D<0)>0.975$, we accept $\hf$. Because we compare
the hypotheses in order to determine when to stop sampling, biasing
the prior should result in requiring less evidence for one hypothesis
before stopping, and more evidence for the other hypothesis.

@@

Like the \Oc{} and \Th{} models, there was an overall monotonic
relationship between rotation and the number of steps taken by the
\Bq{} model (Figure~\ref{fig:response-time-accuracy}). Unlike the
\Th{} model, this relationship existed for individual shapes as well
(e.g., Figure~\ref{fig:response-time-stimulus}).  The \Bq{} model
explained variance in human RTs about as well as the \Th{} model
(Figure~\ref{fig:human-model-scatters}), with a correlation of
\BqTimeCorr{} for the equal prior and \BqpTimeCorr{} for the unequal
prior, and the RT distribution from the \Bq{} model had the same
overall shape as that of people (Figure~\ref{fig:histograms}).\\
\indent The \Bq{} model was quite accurate overall (equal prior:
(\BqAccuracy{}; unequal prior: \BqpAccuracy{}). With the equal prior,
there were \BqNumChance{} stimuli for which it was not above chance;
with the unequal prior, there were \BqpNumChance{}. The correlation
with people's accuracy was \BqAccuracyCorr{} (equal prior) and
\BqpAccuracyCorr{} (unequal prior).\\
\indent Because the \Bq{} model relies on Equation~\ref{eq:lh-ratio}
for its stopping criteria (as opposed to just finding a maximum), the
prior $p(h)$ had an observable effect
(Figure~\ref{fig:response-time-accuracy}). As expected, with just a
small bias of $p(h=0)=0.55$, there was a clear separation in RTs for
``same'' versus ``flipped'' stimuli: because of this bias, the model
needed less evidence before accepting $h=0$ (thus taking less
time). This separation is similar to the trend also observed in human
RTs. The prior also had an effect on accuracy (though this did not
reflect human behavior): the bias towards $h=0$ meant that the model
was more likely to judge a pair as ``same'', thus, accuracy increased
for ``same'' pairs, but decreased for ``flipped'' pairs.

>>>>

We denote the shapes as $X_a$ and $X_b$ and assume $X_b$ is generated
by a transformation of $X_a$, i.e. $X_b=f(X_a, \theta, h)$, where
$\theta$ is a rotation, $\hi$ is the hypothesis that the images depict
the same object, and $\hf$ is the hypothesis that the images depict
mirror-image objects. The posterior probability of each hypothesis
given the observed shapes is then:
$p(h\ \vert\ X_a, X_b) \propto \int p(X_b\ \vert\ X_a, \theta,
h)p(h)p(\theta)\dif\theta$,
where $p(X_b\ \vert\ X_a, \theta, h)$ is the probability that $X_b$
was generated from $X_a$. Because we want to determine which
hypothesis is more likely, the quantity of interest is a posterior
odds ratio
$\mathcal{B}:=p(\hi\ \vert\ X_a, X_b) / p(\hf\ \vert\ X_a, X_b)$ which
(assuming that all rotations are equally likely) is equivalent to:
\begin{equation}
  \mathcal{B} = \frac{\left(\int p(X_b\ \vert\ X_a, \theta, \hi)\dif\theta\right)\cdot{}p_0}{\left(\int p(X_b\ \vert\ X_a, \theta, \hf)\dif\theta\right)\cdot{}p_1},
  \label{eq:odds-ratio}
\end{equation}
where $p_0=p(h=0)$ and $p_1=p(h=1)$, for brevity. If $\mathcal{B} > 1$, then
we accept the hypothesis that the images depict the same object
($\hi$); if $\mathcal{B} < 1$, then we accept the hypothesis that the images
depict flipped objects ($\hf$).

@@

To summarize, we approximate the likelihood term of
Equation~\ref{eq:odds-ratio} using the similarity function defined in
Equation~\ref{eq:similarity}. Because we assume mental rotations are
performed sequentially, this similarity can only be computed for the
actions listed in Equation~\ref{eq:actions}.

@@

In order to approximate Equation~\ref{eq:odds-ratio} using samples of
the similarity function, we must decide \textit{which} places to
sample and \textit{when} stop sampling. The models below differ in how
they make these decisions.

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This gives us the true value of $\theta$, so
Equation~\ref{eq:odds-ratio} becomes a generalized likelihood ratio
test, where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \mathcal{B} = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)\cdot{}p_0}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)\cdot{}p_1}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

While the previous few models all focused on \textit{searching} for
the global maximum, we need only \textit{approximate}
Equation~\ref{eq:odds-ratio}. We hypothesize a model based on the idea
of \textit{active sampling} \cite<e.g.,>{Gureckis:2012gu}: instead of
searching for a maximum, we maintain a probability distribution over
our \textit{estimate} of Equation~\ref{eq:odds-ratio}, and then sample
actions which are expected to improve that estimate.  This strategy
has the benefits that it does not make assumptions about the scale of
the similarity function; and, by choosing to sample places which are
informative, this method implicitly minimizes the amount of rotation.

@@

Assuming independence, we can now write
$p(Z_h)\approx\mathcal{N}(Z_h\ \vert\ \mathbb{E}[Z_h],
\mathbb{V}(Z_h))$,
which gives us a distribution over the likelihood ratio in
Equation~\ref{eq:odds-ratio}:
$p(\mathcal{B})\approx\mathcal{N}(Z_0\ \vert\ \mathbb{E}[Z_0],
\mathbb{V}(Z_0))\cdot{}p_0/\mathcal{N}(Z_1\ \vert\ \mathbb{E}[Z_1],
\mathbb{V}(Z_1))\cdot{}p_1$.
This distribution cannot easily be calculated, but we are only
interested in whether $Z_0>Z_1$ or $Z_1>Z_0$. So, we use
$Z_D=p_0\cdot{}Z_0-p_1\cdot{}Z_1$ and compute
$p(Z_D)\propto\mathcal{N}(p_0\cdot{}\mathbb{E}[Z_0] -
p_1\cdot{}\mathbb{E}[Z_1], p_0^2\cdot{}\mathbb{V}(Z_0) +
p_1^2\cdot{}\mathbb{V}(Z_1))$.
We then sample new observations until we are at least 95\% confident
that $Z_D\neq 0$. In other words, when $p(Z_D<0)<0.025$, we accept
$\hi$, and when $p(Z_D<0)>0.975$, we accept $\hf$. Because we compare
the hypotheses in order to determine when to stop sampling, biasing
the prior should result in requiring less evidence for one hypothesis
before stopping, and more evidence for the other hypothesis.

@@

Like the \Oc{} and \Th{} models, there was an overall monotonic
relationship between rotation and the number of steps taken by the
\Bq{} model (Figure~\ref{fig:response-time-accuracy}). Unlike the
\Th{} model, this relationship existed for individual shapes as well
(e.g., Figure~\ref{fig:response-time-stimulus}).  The \Bq{} model
explained variance in human RTs about as well as the \Th{} model
(Figure~\ref{fig:human-model-scatters}), with a correlation of
\BqTimeCorr{} for the equal prior and \BqpTimeCorr{} for the unequal
prior, and the RT distribution from the \Bq{} model had the same
overall shape as that of people (Figure~\ref{fig:histograms}).\\
\indent The \Bq{} model was quite accurate overall (equal prior:
(\BqAccuracy{}; unequal prior: \BqpAccuracy{}). With the equal prior,
there were \BqNumChance{} stimuli for which it was not above chance;
with the unequal prior, there were \BqpNumChance{}. The correlation
with people's accuracy was \BqAccuracyCorr{} (equal prior) and
\BqpAccuracyCorr{} (unequal prior).\\
\indent Because the \Bq{} model relies on Equation~\ref{eq:odds-ratio}
for its stopping criteria (as opposed to just finding a maximum), the
prior $p(h)$ had an observable effect
(Figure~\ref{fig:response-time-accuracy}). As expected, with just a
small bias of $p(h=0)=0.55$, there was a clear separation in RTs for
``same'' versus ``flipped'' stimuli: because of this bias, the model
needed less evidence before accepting $h=0$ (thus taking less
time). This separation is similar to the trend also observed in human
RTs. The prior also had an effect on accuracy (though this did not
reflect human behavior): the bias towards $h=0$ meant that the model
was more likely to judge a pair as ``same'', thus, accuracy increased
for ``same'' pairs, but decreased for ``flipped'' pairs.
